Section_8_Team1_Lab2_Report

### 2021-08-05 notebooks/Draft_Report.Rmd
updated intro

Summer vacations in the United States are synonymous with sunny, leisurely days spent with loved ones. As Ella Fitzgerald so aptly sings, “Summertime, and the living is easy.” However, travel in the recent peak-vacation months has certainly deviated from this idealistic frame of mind with the onset of the COVID-19 pandemic in early 2020. The global travel industry has experienced sharp declines. Specifically in the United States, “travel spending totaled a mere $679 billion in 2020, an unprecedented 42% annual decline (nearly $500 billion) from 2019,” according to analysis from the U.S. Travel Association. 
As infection and vaccination rates oscillate regionally along with government policies for business operation restrictions in the United States, our team of data scientists will look to answer the following research question in more detail using linear regression techniques:

Do COVID-19 infection rates impact vacation travel distance for U.S. residents in the state of Texas? 
Does income, population density, unemployment rates, and age in Texas counties impact traveling?

We will analyze datasets from June, 2021 pertaining to these variables for Texas counties and create three models to understand if causal relationships exist between the number of long distance trips, population income, population density (rural or urban counties). Our hypothesis is that an there is a causal relationship between """

-->Need to explain distance of travel with bs TRANSPORT data
-->Need to add more to causal theory to justify variable selection

@@

To create our dataframe, we cleaned and filtered data from multiple sources of data and joined based on texas county name or code from June, 2021.

@@

```{r EDA, echo=FALSE, message=FALSE}
#Example Table
df$rural_cat <- ifelse(df$is_rural == 1, "Rural", "Urban")
table1::label(df$density) <- "County Population Density"
table1::label(df$cases) <- "County Covid-19 Cases"
table1::label(df$unemployed) <- "County Unemployment Rate"
table1::label(df$median_income) <- "County Median Income ($)"
table1::label(df$lt17) <- "County Age < 17 years"
table1::label(df$gt65) <- "County Age > 65 years"
# table1::label(df$is_rural) <- "County Type"
table1::table1(~density + cases + unemployed + median_income + lt17 + gt65 | rural_cat, data = df, caption = "Table 1: Texas County Variable Statistics", topclass="Rtable1-zebra", overall = "Total")
```
Below are histograms of variables that did not follow normal distribution and required transformation. Each of these variables was logarithmically transformed.
```{r scatterplots, message=FALSE}
attach(df)
par(mfrow=c(3,2))
hist(cases, main="Histogram of Covid-19 Cases")
hist(deaths, main="Histogram of Covid-19 Deaths")
hist(long_trip, main="Histogram of Distance")
hist(density, main="Histogram of Population Density")
hist(unemployed, main="Histogram of Unemployment")
hist(median_income, main="Histogram of Median Income")

@@

df %>%
  mutate(
    cases = log(cases+1),
    deaths = log(deaths+1),
    long_trip = log(long_trip),
    density = log(density),
    median_income=log(median_income+1),
    unemployed = log(unemployed+1),
    lt17 = log(lt17),
    gt65 = log(gt65)
  ) %>%
  select(long_trip, cases, deaths, median_income, unemployed, density, lt17, gt65) %>%
  GGally::ggpairs(.,
               legend = 1,
               aes(color = County),
               title = "Figure 1: Variable Characterstics",
               upper = list(continuous = wrap("cor", size = 2.5)),
               lower = list(continuous = wrap("points", alpha = 0.5, size=0.3)),
               diag = list(continuous = wrap("densityDiag", alpha = 0.5)),
               columnLabels=c("Distance", "Cases", "Deaths", "Income", "Unemployment", "Pop Density", "Age < 17","Age > 65"))+
               theme(legend.position = "bottom")

@@

To investigate the causal relationships between distance traveled and variables within the dataset, a linear regression was performed in R Studio for three models. The first model includes only our base variables. In the second model, we added additional county characteristics. In the final model, we added in another covid related factor.

Here is the first model, which is limited to study only key variables of interest for measurement based on our hypotheses. Our key input variable is covid cases and key output is number of trips greater than 100 miles. We also chose to add an indicator for rural counties because it could be directly related to trip distance
$$
  Model 1 = f(\text{log(#Long Trips)}) = \beta_0 + \beta_1 g(\text{log(Covid-19 Cases)}) + \beta_2 g(\text{log(Covid-19 Deaths)})  + \beta_3 h(\text{Rural County})
$$ 
The second model expands upon the first model to also include key explanatory variables and covariates that are hypothesized to improve performance of the previous model. Here, we added in median income and density of the county as additional descriptors of the county. 
$$
  Model 2 = f(\text{log(#Long Trips)}) = \beta_0 + \beta_1 g(\text{log(Covid-19 Cases)}) + \beta_2 g(\text{log(Covid-19 Deaths)})  + \beta_3 h(\text{Rural County}) + \beta_4 h(\text{Median Income}) + \beta_5 h(\text{County Density})
$$
Finally in the third model, we added in age demographics as percent under 17/over 65 as another potential source that influences travel midst covid cases due to the nature of how covid affects those at different ages.
$$
  Model 3 = f(\text{log(#Long Trips)}) = \beta_0 + \beta_1 g(\text{log(Covid-19 Cases)}) + \beta_2 g(\text{log(Covid-19 Deaths)})  + \beta_3 h(\text{Rural County}) + \beta_4 h(\text{Median Income}) + \beta_5 h(\text{County Density}) + \beta_6 h(\text{Population under 17}) + \beta_7 h(\text{Population over 65})
$$

@@

#need to change these to per capita data
model1 <- lm(log(long_trip) ~ log(cases+1) + log(deaths+1) + is_rural, data = df) #limited model
model2 <- lm(log(long_trip) ~ log(cases+1) + log(deaths+1) + is_rural + log(median_income+1) + log(density), data = df)
model3 <- lm(log(long_trip) ~ log(cases+1) + log(deaths+1) + is_rural + log(median_income+1) + log(density) + log(lt17) + log(gt65), data = df) #final model
stargazer(model1, model2, model3, type='text', title="Results", align=TRUE)
```

## Discussion
From this output table, we can discern quite a bit of information.

First, we consider the models holistically. Here are some of our observations:
* In model1, all of our variables are extremely significant, though in the opposite direction as expected. We would expect an inverse relationship between covid and travel, though it seems that this model is producing a positive relationship.
* In model2, when county density is factored in, the presence of covid seems to have less positive of an effect on travel and that higher density counties see more travel. 
* In model3 with age factored in, we see that actually the most statistically significant effect on long trips are median income of the household and presence of kids under 17. 
* In terms of R^2 and RSE, we are improving with each model. 
* The addition of our chosen parameters for each following model is also statistically better than the previous, as indicated by the F statistic.

We will also interpret the coefficient values and what these mean in the context of our model. Almost all of our variables needed a log transform, including the output, making this a log-log regression. These regressions are most intuitive to understand in terms of percent increases. 
* First to note is our persistently significant constant in each model. Regardless of any variables, we are getting between ~300 (as with model3) and over 8000 (as with model2) trips over 100 miles. This is calculated as e^(Constant). 
* In model1, for every percent increase in both cases and deaths, we are getting almost half a percent increase in trips. The positive relationship doesnt make much sense, so we will look to the other models.
* In model2, we see that density has a more positive relationship with trips, with every percent increase in county density modeled to a half percent increase in trips. Also the positive relationship between covid cases/ deaths and trips has decreased to resulting in 0.15 and 0.25% increase in trips respectively.
* In model3, covid cases and deaths become insignificant. The coefficients have become slightly negative as in increase in cases and deaths decreases trips, which is in line with our original hypothesis, but is not significant. Instead, for every percent median income of the county increases by, the amount of long trips decrease by 0.3%. This could indicate that more affluent counties trust science and err on the safe side in the midst of covid. Finally, in our 3rd model, counties with more people under 17 years of age are more likely to take long trips. Due to the time frame of our analysis, this also makes sense because june is the start of summer break when many families like to take vacations together.

@@

DELTE*** Not CLM since >100 samples. Large sample assumptions = IID and unique BLP (which is related to colinearity...)

@@

The goal of this analysis was to determine if there existed a causal relationship between distance traveled and Covid-19 infection rates for residents of the state of Texas. Based upon the linear regression analysis above, model 2 would indicate there may be a causal relationship between distance traveled (outcome variable) and the variables found within that model, though with all variables considered, our results do not indicate vacation and covid cases are necessarily related.

For future analysis, it is recommended that Covid-19 policy adherence data per county be incorporated into the model to increase the accuracy. While the Governor of Texas issued detailed protocols or mandates for the state of Texas, the degree to which each county (and thus the individual residents) complied would be useful data to have if such data were to exist. Additionally, incorporating vaccination rates could aid in development of a model with even higher accuracy. Generally, vaccinations could increase confidence among individuals interested in traveling, especially if they have been in long periods of quarantine or government mandated lockdowns. Last, it is highly recommended that any model generated in the future continue to use the latest data available given the oscillation in vaccination rates, government policy, and new or known variants of the Covid-19 virus.

>>>>

Summer vacations in the United States are synonymous with sunny, leisurely days spent with loved ones. As Ella Fitzgerald so aptly sings, “Summertime, and the living is easy.” However, travel in the recent peak-vacation months has certainly deviated from this idealistic frame of mind with the onset of the COVID-19 pandemic in early 2020. The global travel industry has experienced sharp declines in Revenue. Specifically in the United States, “travel spending totaled a mere $679 billion in 2020, an unprecedented 42% annual decline (nearly $500 billion) from 2019,” according to analysis from the U.S. Travel Association. Our hypothesis as to why this decline has occurred is due directly to the presence of the Covid-19 virus and the number of cases. As virus infection oscillate regionally in the United States, our team of data scientists will look to answer the following primary research question using linear regression statistical techniques:

Do COVID-19 cases impact vacation travel trips for Texas for June, 2021?

Our primary hypothesis is that there is a causal relationship between the number of Covid-19 cases and the number of vacation trips; when Covid-19 cases increase, the number of vacation travel trips decrease. To investigate, a dataset comprising of Texas county data will be analyzed. Based on the data, our team will also investigate the following secondary question:

Do income, population density, age, and/or county type (rural or urban) per county impact traveling for Texas residents?

For the analysis, we will create three models to understand if causal relationships exist between the outcome variable, the number of long distance trips, and the predictor variables per Texas county: income, population density, age, Covid-19 cases and deaths . We will use an indicator variable, rural counties, to compare performance of our models for rural or urban counties.

@@

To create our dataframe, we cleaned and filtered data from multiple sources of data and joined based on texas county name or code.

@@

```{r EDA, echo=FALSE, message=FALSE}
#Example Table
df$rural_cat <- ifelse(df$is_rural == 1, "Rural", "Urban")
table1::label(df$density) <- "County Population Density"
table1::label(df$cases) <- "County Covid-19 Cases"
table1::label(df$unemp_rate) <- "County Unemployment Rate"
table1::label(df$median_income) <- "County Median Income ($)"
table1::label(df$pct_lt17) <- "County Age < 17 years (%)"
table1::label(df$pct_gt65) <- "County Age > 65 years (%)"
# table1::label(df$is_rural) <- "County Type"
table1::table1(~density + cases + unemp_rate + median_income + pct_lt17 + pct_gt65 | rural_cat, data = df, caption = "Table 1: Texas County Variable Statistics", topclass="Rtable1-zebra", overall = "Total")
```
Below are histograms of variables that did not follow normal distribution and required transformation. Each of these variables was logarithmically transformed.
```{r scatterplots, message=FALSE}
attach(df)
par(mfrow=c(3,2))
hist(cases, main="Histogram of Covid-19 Cases")
hist(deaths, main="Histogram of Covid-19 Deaths")
hist(long_trip, main="Histogram of Trip Count")
hist(density, main="Histogram of Population Density")
hist(unemp_rate, main="Histogram of Unemployment Rate")
hist(log(median_income), main="Histogram of Median Income")
hist(median_income, main="Histogram of Median Income")
summary(df$median_income)
summary(df$long_trip)

@@

df %>%
  mutate(
    cases = log(cases+1),
    deaths = log(deaths+1),
    long_trip = log(long_trip),
    density = log(density),
    median_income=log(median_income+1),
    unemp_rate = log(unemp_rate+1)
  ) %>%
  select(long_trip, cases, deaths, median_income, unemp_rate, density, pct_lt17, pct_gt65) %>%
  GGally::ggpairs(.,
               legend = 1,
               aes(color = County),
               title = "Figure 1: Variable Characterstics",
               upper = list(continuous = wrap("cor", size = 2.5)),
               lower = list(continuous = wrap("points", alpha = 0.5, size=0.3)),
               diag = list(continuous = wrap("densityDiag", alpha = 0.5)),
               columnLabels=c("Distance", "Cases", "Deaths", "Income", "Unemployment", "Pop Density", "Age < 17","Age > 65"))+
               theme(legend.position = "bottom")

@@

To investigate the causal relationships between distance traveled and variables within the dataset, a linear regression was performed in R Studio for three models. The first model is limited to study only key variables of interest for measurement. The second model expands upon the first model to also include key explanatory variables and covariates that are hypothesized to improve performance of the previous model. The third model is all-inclusive of the variables available to the team at the time of data gathering. Details of each model analyzed can be found below.

$$
  Model 1 = f(\text{log(Number of Long Trips)}) = \beta_0 + \beta_1 g(\text{log(Covid-19 Cases)})  + \beta_2 h(\text{Rural County}) + \beta_3 h(\text{Median Income})
$$

$$
  Model 2 = f(\text{log(Number of Long Trips)}) = \beta_0 + \beta_1 g(\text{log(Covid-19 Cases)})  + \beta_2 h(\text{Rural County}) + \beta_3 h(\text{Median Income}) + \beta_4 h(\text{Unemployment Rate}) + \beta_5 h(\text{log(Population Density})
$$

$$
  Model 3 = f(\text{log(Number of Long Trips)}) = \beta_0 + \beta_1 g(\text{log(Covid-19 Cases)})  + \beta_2 h(\text{Rural County}) + \beta_3 h(\text{Median Income})
  + \beta_4 h(\text{Unemployment Rate}) + \beta_5 h(\text{log(Population Density}) + \beta_6 h(\text{log(% Age < 17}) + \beta_7 h(\text{log(% Age < 17})
$$

@@

#need to change these to per capita data
model1 <- lm(log(long_trip) ~ log(cases+1) + log(deaths+1) + is_rural + log(median_income+1), data = df) #limited model
summary(model1)
model1b <- lm(log(long_trip) ~ log(cases+1) + log(deaths+1) + is_rural + median_income, data = df) #limited model
summary(model1b)
summary(df$median_income)

model2 <- lm(log(long_trip) ~ log(cases+1) + log(deaths+1) + is_rural + log(median_income) + unemp_rate + log(density), data = df)
model3 <- lm(log(long_trip) ~ log(cases+1) + log(deaths+1) + is_rural + log(median_income)+ unemp_rate + log(density) + pct_lt17 + pct_gt65, data = df) #final model
stargazer(model1, model1b, model2, model3, type='text', title="Results", align=TRUE)
```
Additionally, to indicate if any of the models provide a better fit to the dataset than a model that contains no independent variables, an F test was performed. The results for each model are shown below:
```{r F Test}
#Run F test to show statistical significance of additional variables
anova(model1, model2, model3, test = "F")
```
From the F-test,  we can see the the addition of the variables into model 2 and model 3 do contribute signficant information to our model based off of the derived p-values.

Last, the significance of individual regression coefficients was tested using robust standard errors in R studio for model 3. The results are shown below: 
```{r coeftest}
#Run coeftest to show statistical significance of all variables in the model with robust standard errors
coeftest(model2, vcov = vcovHC(model3, type="HC1"))
```
From the t-test, we can see that Covid-19 cases and median income are statistically significant. Additionally, the percentage of adults greather than 65, the population density, and the Covid-19 death variables are even more statistically significant.

## Discussion

@@

@@

The goal of this analysis was to determine if there existed a causal relationship between distance traveled and Covid-19 infection rates for residents of the state of Texas. Based upon the linear regression analysis above, model 2 would indicate there may be a causal relationship between distance traveled (outcome variable) and the variables found within that model.

For future analysis, it is recommended that Covid-19 policy adherence data per county be incorporated into the model to increase the accuracy. While the Govenor of Texas issued detailed protocols or mandates for the state of Texas, the degree to which each county (and thus the individual residents) complied would be useful data to have if such data were to exist. Additionally, incorporating vaccination rates could aid in development of a model with even higher accuracy. Generally, vaccinations could increase confidence among individuals interested in traveling, especially if they have been in long periods of quarantine or government mandated lockdowns. Last, it is highly recommended that any model generated in the future continue to use the latest data available given the oscillation in vaccination rates, government policy, and new or known variants of the Covid-19 virus.
